\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-04-19 doc/incertae_sedis.tex
line wrapping

%% LyX 2.3.0 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=3cm}
\setlength{\parskip}{6pt}
\setlength{\parindent}{0pt}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[authoryear]{natbib}

@@

\usepackage{babel}
\begin{document}
\global\long\def\taxonomy{\mbox{\ensuremath{\mathbb{T}}}}

@@

@@

\author{Benjamin D. Redelings and Mark T. Holder}
\maketitle

@@

While some taxonomies may indicate hybrid taxa, the supertree approach
described here does not consider any hybridization. Thus the taxonomic
hierarchy can be converted to a tree. While not all taxonomists believe
that named taxa should correspond to clades \citep[see, for example,][]{HorandlS2010},
the priniciples of phylogenetic classification are used by such a
large majority of practicing taxonomists, that we have chosen to treat
the taxonomic tree that mirrors the hierarchy of the taxonomy as an
estimate of phylogenetic relationships. We refer to labels of taxonomy
nodes as ``taxon names'', and assume that taxon names and taxonomy
nodes have a one-to-one correspondence. Note that the taxonomy may
also contain out-degree 1 nodes. These nodes correspond to monotypic
taxa which contain a single child of lower taxonomic rank.

@@

Because of the large number (over 2.6 million) of tips in the full
tree, the supertree method of \citet{redelings2017supertree} uses
an approximation approach that relies on decomposition of the full
problem into subproblems. The decomposition partitioning the list
of taxonomy splits $S(\taxonomy_{p})$ into two lists: $\mathcal{Z}(\taxonomy_{p}),$the
list of splits from the taxonomy tree which are compatible with every
rooted split in the phylogenetic inputs, and $\mathcal{Y}(\taxonomy_{p})$,
the list of taxonomic splits that conflict with at least one phylogenetic
input split. Conceptually the approximate algorithm works by greedily
adding all compatible splits in altered ranked list of splits: $\mathcal{S}^{\prime}=\mathcal{Z}(\taxonomy_{p})+S(T_{1})+S(T_{2})+\ldots+S(T_{n})+\mathcal{Y}(\taxonomy_{p})$.

@@

1.1 Some examples in OTT This paragraph pulled from intro to otcetera/doc/handling-incertae-sedis.pdf

The problem is becoming more acute because NCBI is putting a larger number of taxa into groups that are marked as “unclassified.” For example, when OTT 2.9 was created NCBI’s clasification of the bird family Sylvidae included a group five genera that were placed within “[http://www.ncbi.nlm.nih.gov/Taxonomy/Browser/wwwtax.cgi?mode=Undef&id=36270&lvl=3&lin=f&keep=1&srchmode=1&unlock||unclassified Sylviidae].” This includes the genus Regulus. Thus in OTT 2.9 Regulus (ott ID = 3599326) is placed inside the family Silvidae (OTT ID = 259942). The “unclassified Sylviidae” does not appear in OTT; instead Regulus is flagged as “unclassified,sibling higher” and all of the species within Regulus are flagged as “unclassified inherited.”

@@

Flags. The Open Tree’s reference taxonomy is produced by a tool called smasher that notices hints that a taxon is incertae sedis and labels that taxon with one of five flagssee https://github.com/OpenTreeOfLife/reference-taxonomy/wiki/Taxon-flags.This document is not concerned with the flagging system per se, so “incertae sedis” will be used here to refer to all of the flags that denote taxa with uncertain placement.. While a taxonomist may intend to use incertae sedis to indicate a limited number of possible positions for a taxon to go in the taxonomy, we do not retain any such details.

@@

Note that, for the purposes of the supertree method, all of the information
from the taxonomic tree is contained in the set of rooted splits.
Thus, one could imagine replacing that taxonomy, $\taxonomy_{p}$,
with a set of trees $\mathcal{R}(\taxonomy_{p})$ each of which has
a single internal edge such that each edge in $\taxonomy_{p}$is included
in one tree in $\mathcal{R}(\taxonomy_{p})$. If the order of trees
is set to be (with some abuse of notation): $\mathcal{R}(\taxonomy_{p})=\mathcal{Z}(\taxonomy_{p})+\mathcal{Y}(\taxonomy_{p})$,
then using $\mathcal{R}(\taxonomy_{p})$ in place of $\taxonomy_{p}$in
the supertree algorithm would produce the same output.

@@

\paragraph{Information content of the \emph{incertae sedis }annotation}

@@

DO WE HAVE SOME NUMBERS ON THIS (PRESUMABLY FROM JUST RUNNING THE
ONLY PROPINQUITY WITH A DIFFERENT SET OF PRUNE FLAGS)?

@@

We seek a semantics for supertrees with \emph{incertae sedis }taxa
that satisfy the following properties:
\begin{enumerate}
\item An \emph{incertae sedis} node may intrude into its siblings and their
descendants.
\item Two \emph{incertae sedis} siblings may not be interdigitated.
\item Any placement $T\to T^{*}$ of a tree $T$ with \emph{incertae sedis}
taxa is also accessible from a tree $T^{\prime}$ that is created
from $T$ by taking a taxon, attaching it to an ancestor, and marking
it \emph{incertae sedis}.%
\begin{comment}
MTH: I don't understand this
\end{comment}
\item The semantics of an \emph{incertae sedis} node does not depend on
assigned ranks, but only on the tree.
\item Semantics is based on deriving a split for each branch of the tree.
\end{enumerate}
Here we focus on a split-based semantics instead of a semantics based
on apply of a series of edit operations to the tree. The split-based
semantics allows us to seek trees that are simultaneously consistent
with a set of split-constraints. The use of a split-based semantics
does impose some limitations. These come primarily because in the
edit-based semantics, after we place an \emph{incertae sedis} taxon,
it is no loner \emph{incertae sedis}, whereas with the split-based
semantics taxa are either always \emph{incertae sedis} or never \emph{incertae
sedis}.

This has an effect, for example, when we consider two incertae sedis
siblings $A$ and $B$. If we can place $A$ within $B$, and $B$
within $A$, then a split-based semantics allows us to freely inter-digitate
$A$ and $B$. If we do not want to allow $A$ and $B$ to be inter-digitated,
then we cannot both $A$ within $B$ and $B$ within $A$. In addition,
we can imagine that a family $A$ is incertae sedis within a kingdom.
In an edit-based semantics, we could place $A$ as a sibling to a
genus $B$, mark $A$ as non-\emph{incertae-sedis}, and then place
$B$ within $A$. In a split-based semantics, we must disallow either
$B$ within $A$ or $A$ within $B$ if we want to disallow interdigitation.

As a result of this, we cannot satisfy all five of these criteria.
Thus, we describe two alternative split-based semantics. The non-interdigitation
semantics gets \# (1,2,4,5), and the interdigitation semantics gets
(1,3,4,5). This makes the point that each semantics results from different
choices about which criteria to satisfy.

\subsection{Split-based semantics for \emph{incertae sedis} taxa}

@@

\begin{align*}
\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup[\mathcal{I}^{\prime}(s)|s\in siblings(n)].
\end{align*}

@@

We may then define the taxa excluded from $n$ as:

\begin{align*}
\mathcal{E}(n) & =\begin{cases}
\left[\mathcal{L}-\mathcal{I}(parent(n))\right]\cup[\mathcal{I}(s,n)|s\in siblings(n)] & \text{if }n\in\mathcal{U}\\
\mathcal{E}(parent(n))\cup[\mathcal{I}(s,n)|s\in siblings(n)] & \text{otherwise.}
\end{cases}
\end{align*}
I think this works... but need to double-check.

@@

However, in the \emph{incertae sedis} framework, it is possible for
one name to apply to multiple nodes. For example, in Figure \ref{fig:One-name-can}b,
the name $A$ can apply to nodes $x$ and $y$. Here the name $A$
corresponds to the split $A1\,A2\,|\bullet B\,C$, leaving out $D$
since it is \emph{incertae sedis}. The two nodes $x$ and $y$ display
the splits $A1\,A2\,|\bullet B\,C\,D$ and $A1\,A2\,D|\bullet B\,C$
respectively, and both of these splits display the split $A1\,A2\,|\bullet B\,C$,
so the name $A$ can apply to both $x$ and $y$. This cannot happen
without\emph{ incertae sedis} taxa except at monotypic nodes. When
faced with a choice about where to place a name, our solution is to
find the most tip-ward node where the name can apply and attach the
name to this node. This corresponds to a conservative choice about
whether \emph{incertae sedis} taxa should be added to an existing
taxon.
\begin{figure}
\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/name1nodes2/tax-shaded}

}\hfill{}\subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/name1nodes2/solution-shaded}

\caption{\label{fig:One-name-can}One name can be consistent with multiple
nodes.}
\end{figure}

It is also possible for multiple names to apply to a single node.
For example, in Figure \ref{fig:Multiple-names-can}a, the taxon $A$
corresponds to the split $B1\,B2\,C\,|\bullet\,Y$, and its child
taxon $B$ corresponds to the split $B1\,B2\,|\bullet\,Y$. The edge
leading to $A$ is consistent with the split for $B$, but the name
$B$ is applied to the node with the smallest include group. However,
in the solution tree (Fig. \ref{fig:Multiple-names-can}b), there
is is only node for both names $A$ and $B$ to apply to. In this
case, we solve this problem by introducing a monotypic parent, and
applying $A$ to the newly created parent node (Fig. \ref{fig:Multiple-names-can}c).
\begin{figure}
\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/names2nodes1/tax-shaded}

}\hfill{}\subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/names2nodes1/synth-shaded}

}\hfill{}\subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/names2nodes1/solution2-shaded}

\caption{\label{fig:Multiple-names-can}Multiple names can apply to a single
node.}

@@

\subsection{\label{subsec:Placement}Placement}

After we have attached taxon names to the synthesis tree, we would
like to interpret the position of these names in terms of placing
\emph{incertae sedis }taxa in a revised taxonomy. This would allow
us to interpret the synthesis tree as saying that phylogenetic information
has (for example) placed genus A within family B, or perhaps outside
of all named families. The simplest approach to placement involves
noting whenever a taxon $B$ is a descendant of a taxon $A$ on the
named synthesis tree but not the taxonomy tree. For example, in Figure
\ref{fig:Handling-incertae-sedis}, taxon $C$ is a descendant of
taxon $A$ in the synthetic tree (Fig. \ref{fig:Handling-incertae-sedis}d),
but not the taxonomy (Fig. \ref{fig:Handling-incertae-sedis}b). Thus,
we could say that the synthetic tree places $C$ within $A$. Furthermore,
the synthetic tree additionally places $C$ within $AB$.

A slightly more general case is shown in Figure \ref{fig:Broken-incertae-sedis}.
Here taxon $C1$ is an descendant of $A$ in the synthesis tree, but
not the taxonomy. Thus $C1$ is placed within $A$. We note that the
most recent common ancestor (MRCA) of $A$ and $C1$ in the taxonomy
is the root node, and the path to the MRCA from $A$ is $A\to AB\to root$,
while the path to the MRCA from $C1$ is $C1\to C\to root$. Here
the additional node $AB$ on the path to the MRCA from $A$ indicates
that $C1$ is placed within $AB$ as well as $A$. In contrast, the
additional node $C$ on the path to the MRCA indicates that $C1$
is a broken \emph{incertae sedis} taxon. Thus we may consider an algorithm
that, for each taxon $B$ on the synthesis tree, finds the closest
ancestral taxon $A$, and checks if $A$ is an ancestor of $B$ on
the taxonomy. If not, we compute the paths $A\to A_{1}\to\ldots\to MRCA(A,B)$
and $B\to B_{1}\to\ldots\to MRCA(A,B)$.

However, this assumes that \emph{incertae sedis} taxa are always placed
within their siblings, or descendants of their siblings. As such,
it does not handle cases such as Figure \ref{fig:Placing-one-incertae},
in which $I$ is first placed as sibling to $AI$, and then $AI$
is placed within $I$. Note that $I$ is not a sibling of $AI$ or
a descendant of a sibling. Furthermore, the MRCA of $I$ and $AI$
is the root, and the path from $AI$ to the root includes $A$, but
$A$ is not broken. We suggest that this conundrum may be resolved
by first processing the placement of $I$ within $A$ by modifying
the taxonomy to reflect this placement\footnote{Does this actually work? Some placements break taxa.}.
When we consider the placement of $AI$ within $I$ on this modified
taxonomy, the MRCA of $I$ and $AI$ is now $A$ instead of the root,
and the correct interpretation results. We suggest that by walking
the tree in a pre-order fashion and processing shallowing placements
first, the previous approach of finding the MRCA and labelling intermediate
nodes as either additional placements or broken incertae sedis (or
incertae\_sedis\_inherited) taxa holds. Such an approach should also
handle cases containing nested incertae sedis taxa, as in Figure \ref{fig:Placing-one-incertae}.%
\begin{comment}
One question is whether this approach is able to handle nested incertae
sedis taxa.

@@

Each input tree can relate to an \emph{incertae sedis} taxon $(A,B,C)D$
in a number of ways
\begin{itemize}
\item it could resolve $A$, $B$, $C$, or $D$.
\item it could place $D$ on a degree-2 (=out-degree-1) node that bisects
a branch
\item it could place a descendant taxon of $D$ 
\item it could place a descendant taxon of $D$ in a \emph{different place}
than another input tree.
\item it could place children of $D$ in multiple places, thus conflicting
with the branch. If the \emph{incertae sedis} taxon $(A,B,C)$ is
broken, then $A$, $B$ and $C$ become \emph{incertae sedis} clades
in their own right, that may attach separately, except that they .
This is because none of $A$, $B$, or $C$ is in the exclude set
of the siblings of $D$.
\end{itemize}
\end{comment}

@@

When the synthesis tree conflicts with a taxonomy node, we say that
the taxon $B$ at that node is a broken taxon. Broken taxa have two
main effects, both of which are negative. First, the name $B$ of
the broken taxon is removed from the synthesis tree. Second, the conflicting
edge for taxon $B$ is not included in the synthesis tree. This means
that any children of $B$ that are taxonomy-only will not be placed
with the children of $B$ that are mentioned in input phylogenies.
Instead the taxonomy-only children of a broken taxon move towards
the root of the synthesis tree and attach at the first higher-ranked
taxon that is an ancestor of $B$ but is not broken.

We therefore seek a synthesis method that can correctly introduce
\emph{incertae sedis} taxa into containing taxa without breaking the
containing taxa. We change the semantics of names to imply, not the
exclusion of all non-included taxa, but only some non-included taxa,
as described in section \ref{sec:Semantics-of-incertae}. As a result
of this change in semantics, placing an IS taxon $A$ within a sister
taxon $B$ no longer results in conflict with $B$. This allows us
to retain the split for $B$ within the synthesis tree, so that taxonomy-only
children of $B$ are correctly grouped with their siblings that are
referenced by the input trees. We may then retain the name for the
no-longer-broken taxon. Finally, we are then able to stop filtering
\emph{incertae sedis} taxa, so that they appear in the synthesis tree.
Thus, the synthesis tree is able to represent substantially more species,
without suffering the loss of taxa and the loss of structure.

@@

{[}\textbf{BDR:} \emph{maybe we should merge this with Figure. \ref{fig:Broken-incertae-sedis}.
Although the point of figure \ref{fig:placement-example1} is to talk
about sub-problem decomposition as well as what the result looks like.}{]}

The addition of \emph{incertae sedis} taxa allows new types of conflict
between input trees. For example, different input trees might place
an incertae sedis taxon in conflicting locations. This is illustrated
in Figure \ref{fig:placement-example1}, where the IS taxon (ott7,ott6)ott10
is placed as sister to ott1 by phylogeny $\tau_{1}$ and as sister
ott3 by phylogeny $\tau_{2}$.

@@

\subsection{Case 1: An \emph{incertae sedis} clade}

When a clade is marked as \emph{incertae sedis}, we need to consider
two cases. In the first case, the clade may be placed within a sister
clade intact (Figure \ref{fig:Handling-incertae-sedis}). However,
if the clade is not monophyletic in the synthesis tree, then we allow
the members of the clade to be placed separately (Figure \ref{fig:Broken-incertae-sedis}).

@@

\begin{figure}
\hfill{}\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/tree1-shaded}

@@

\subsection{Case 2: A nested \emph{incertae sedis} clade}

@@

\emph{}
\begin{figure}
\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}

@@

\section{Handling \emph{incertae sedis} taxa in the propinquity pipeline}

@@

One current problem is that well-known taxa like Fungi or Mammalia
tend to have a very large number of incertae sedis children, making
browsing in the tree viewer difficult. This can happen when, for example,
fossils or other hard-to-place taxa get classified only to the level
of these well-known nodes and no further. This leads to a situation
where well-known taxa serve as a dumping ground for unplaced taxa.

@@

The presence of \emph{incertae sedis }taxa poses a problem to sub-problem
decomposition, since taxonomy edges no longer completely separate
subproblems. Instead, \emph{incertae sedis} taxa may attach on either
side of a taxonomy edge. We seek to place \emph{incertae sedis} taxa
into subproblems in such a way that the subproblem solver can perform
the placement inside the subproblem. This approach postpones handling
of conflict in \emph{incertae sedis} taxa to the subproblem solver,
where the problem is well formulated in terms of splits. However,
it does have the effect of creating larger subproblems.

>>>>

\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=3cm}
\setlength{\parskip}{6pt}
\setlength{\parindent}{0pt}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[authoryear]{natbib}

@@

\usepackage{babel}
\usepackage{xspace}
\usepackage{color}
\begin{document}

\global\long\def\taxonomy{\mbox{\ensuremath{\mathbb{T}}}}

@@

\newcommand{\mthcomment}[1]{{\color{red} #1}\xspace}

@@

\author{Benjamin D. Redelings and Mark T. Holder} maketitle

@@

While some taxonomies may indicate hybrid taxa, the supertree approach
described here does not consider any hybridization. Thus the taxonomic
hierarchy can be converted to a tree. While not all taxonomists
believe that named taxa should correspond to clades \citep[see, for
example,][]{HorandlS2010}, the priniciples of phylogenetic
classification are used by such a large majority of practicing
taxonomists, that we have chosen to treat the taxonomic tree that
mirrors the hierarchy of the taxonomy as an estimate of phylogenetic
relationships. We refer to labels of taxonomy nodes as ``taxon
names'', and assume that taxon names and taxonomy nodes have a one-to-
one correspondence. Note that the taxonomy may also contain out-degree
1 nodes. These nodes correspond to monotypic taxa which contain a
single child of lower taxonomic rank.

@@

Because of the large number (over 2.6 million) of tips in the full
tree, the supertree method of \citet{redelings2017supertree} uses an
approximation approach that relies on decomposition of the full
problem into subproblems. The decomposition partitioning the list of
taxonomy splits $S(\taxonomy_{p})$ into two lists:
$\mathcal{Z}(\taxonomy_{p}),$the list of splits from the taxonomy tree
which are compatible with every rooted split in the phylogenetic
inputs, and $\mathcal{Y}(\taxonomy_{p})$, the list of taxonomic splits
that conflict with at least one phylogenetic input split. Conceptually
the approximate algorithm works by greedily adding all compatible
splits in altered ranked list of splits:
$\mathcal{S}^{\prime}=\mathcal{Z}(\taxonomy_{p})+S(T_{1})+S(T_{2})+\l
dots+S(T_{n})+\mathcal{Y}(\taxonomy_{p})$.

@@

1.1 Some examples in OTT This paragraph pulled from intro to
otcetera/doc /handling-incertae-sedis.pdf

The problem is becoming more acute because NCBI is putting a larger
number of taxa into groups that are marked as “unclassified.” For
example, when OTT 2.9 was created NCBI’s clasification of the bird
family Sylvidae included a group five genera that were placed within
“[http://www.ncbi.nlm.nih.gov/Taxonomy/Brows er/wwwtax.cgi?mode=Undef&
id=36270&lvl=3&lin=f&keep=1&srchmode=1&unlock||unclassi fied
Sylviidae].” This includes the genus Regulus. Thus in OTT 2.9 Regulus
(ott ID = 3599326) is placed inside the family Silvidae (OTT ID =
259942). The “unclassified Sylviidae” does not appear in OTT; instead
Regulus is flagged as “unclassified,sibling higher” and all of the
species within Regulus are flagged as “unclassified inherited.”

@@

Flags. The Open Tree’s reference taxonomy is produced by a tool called
smasher that notices hints that a taxon is incertae sedis and labels
that taxon with one of five flagssee https://github.com/OpenTreeOfLife
/reference-taxonomy/wiki /Taxon-flags.This document is not concerned
with the flagging system per se, so “incertae sedis” will be used here
to refer to all of the flags that denote taxa with uncertain
placement.. While a taxonomist may intend to use incertae sedis to
indicate a limited number of possible positions for a taxon to go in
the taxonomy, we do not retain any such details.

@@

Note that, for the purposes of the supertree method, all of the
information from the taxonomic tree is contained in the set of rooted
splits. Thus, one could imagine replacing that taxonomy,
$\taxonomy_{p}$, with a set of trees $\mathcal{R}(\taxonomy_{p})$ each
of which has a single internal edge such that each edge in
$\taxonomy_{p}$is included in one tree in
$\mathcal{R}(\taxonomy_{p})$. If the order of trees is set to be (with
some abuse of notation):
$\mathcal{R}(\taxonomy_{p})=\mathcal{Z}(\taxonomy_{p}) + 
\mathcal{Y}(\taxonomy_{p})$, then using $\mathcal{R}(\taxonomy_{p})$ in
place of $\taxonomy_{p}$in the supertree algorithm would produce the
same output.

@@

\paragraph{Information content of the emph{incertae sedis }annotation}

@@

\mthcomment{ Do we have some numbers on this (presumably from just
running the only propinquity with a different set of prune flags)? }

@@

We seek a semantics for supertrees with \emph{incertae sedis }taxa
that satisfy the following properties:
\begin{enumerate}
    \item An
\emph{incertae sedis} node may intrude into its siblings and their
descendants.
\item Two \emph{incertae sedis} siblings may not be
interdigitated.
\item Any placement $T\to T^{*}$ of a tree $T$ with
\emph{incertae sedis} taxa is also accessible from a tree $T^{\prime}$
that is created from $T$ by taking a taxon, attaching it to an
ancestor, and marking it \emph{incertae sedis}.
\begin{comment} MTH:
I don't understand this \end{comment} \item The semantics of an
\emph{incertae sedis} node does not depend on assigned ranks, but only
on the tree.
\item Semantics is based on deriving a split for each
branch of the tree.
\end{enumerate}
Here we focus on a split-based
semantics instead of a semantics based on apply of a series of edit
operations to the tree. The split-based semantics allows us to seek
trees that are simultaneously consistent with a set of split-
constraints. The use of a split-based semantics does impose some
limitations. These come primarily because in the edit-based semantics,
after we place an \emph{incertae sedis} taxon, it is no loner
\emph{incertae sedis}, whereas with the split-based semantics taxa are
either always \emph{incertae sedis} or never \emph{incertae sedis}.

This has an effect, for example, when we consider two incertae sedis
siblings $A$ and $B$. If we can place $A$ within $B$, and $B$ within
$A$, then a split- based semantics allows us to freely inter-digitate
$A$ and $B$. If we do not want to allow $A$ and $B$ to be inter-
digitated, then we cannot both $A$ within $B$ and $B$ within $A$. In
addition, we can imagine that a family $A$ is incertae sedis within a
kingdom. In an edit-based semantics, we could place $A$ as a sibling
to a genus $B$, mark $A$ as non-\emph{incertae-sedis}, and then place
$B$ within $A$. In a split-based semantics, we must disallow either
$B$ within $A$ or $A$ within $B$ if we want to disallow
interdigitation.

As a result of this, we cannot satisfy all five of these criteria.
Thus, we describe two alternative split-based semantics. The non-
interdigitation semantics gets \# (1,2,4,5), and the interdigitation
semantics gets (1,3,4,5). This makes the point that each semantics
results from different choices about which criteria to satisfy.

\subsection{Split-based semantics for emph{incertae sedis} taxa}

@@

\begin{align*} 
\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I}(m,n)\big|m\in siblings(n)\right].\label{eq:exclude-set-formula-1}
\end{align*}

@@

We may then define the taxa excluded from $n$ as:
\begin{align*}
\mathcal{E}(n) & =\begin{cases}
\left[\mathcal{L}-\mathcal{I}(parent(n))\right]\cup[\mathcal{I}(s,n)|s\in siblings(n)] & \text{if }n\in\mathcal{U}\\
\mathcal{E}(parent(n))\cup[\mathcal{I}(s,n)|s\in siblings(n)] & \text{otherwise.}
\end{cases}
\end{align*}
I think this works... but
need to double-check.

@@

However, in the \emph{incertae sedis} framework, it is possible for
one name to apply to multiple nodes. For example, in Figure \ref{fig
:One-name-can}b, the name $A$ can apply to nodes $x$ and $y$. Here the
name $A$ corresponds to the split $A1\,A2\,|\bullet B\,C$, leaving out
$D$ since it is \emph{incertae sedis}. The two nodes $x$ and $y$
display the splits $A1\,A2\,|\bullet B\,C\,D$ and $A1\,A2\,D|\bullet
B\,C$ respectively, and both of these splits display the split
$A1\,A2\,|\bullet B\,C$, so the name $A$ can apply to both $x$ and
$y$. This cannot happen without\emph{ incertae sedis} taxa except at
monotypic nodes. When faced with a choice about where to place a name,
our solution is to find the most tip-ward node where the name can
apply and attach the name to this node. This corresponds to a
conservative choice about whether \emph{incertae sedis} taxa should be
added to an existing taxon.
\begin{figure}
\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/name1nodes2/tax-shaded}}
\hfill{}\subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/name1nodes2/solution-shaded}}
\hfill{}
\caption{\label{fig:One-name-can}One name can be consistent with multiple nodes.}
\end{figure}

It is also possible for multiple names to apply to a single node. For
example, in Figure \ref{fig:Multiple-names-can}a, the taxon $A$
corresponds to the split $B1\,B2\,C\,|\bullet\,Y$, and its child taxon
$B$ corresponds to the split $B1\,B2\,|\bullet\,Y$. The edge leading
to $A$ is consistent with the split for $B$, but the name $B$ is
applied to the node with the smallest include group. However, in the
solution tree (Fig. \ref{fig:Multiple-names-can}b), there is is only
node for both names $A$ and $B$ to apply to. In this case, we solve
this problem by introducing a monotypic parent, and applying $A$ to
the newly created parent node (Fig. \ref{fig:Multiple-names-can}c).
\begin{figure}
\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/names2nodes1/tax-shaded}}
\hfill{}subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/names2nodes1/synth-shaded}}
\hfill{}subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/names2nodes1/solution2-shaded}}
\hfill{}
\caption{\label{fig:Multiple-names-can}Multiple names can apply to a single node.}
\end{figure}

@@

\subsection{label{subsec:Placement}Placement}

After we have attached taxon names to the synthesis tree, we would
like to interpret the position of these names in terms of placing
\emph{incertae sedis }taxa in a revised taxonomy. This would allow us
to interpret the synthesis tree as saying that phylogenetic
information has (for example) placed genus A within family B, or
perhaps outside of all named families. The simplest approach to
placement involves noting whenever a taxon $B$ is a descendant of a
taxon $A$ on the named synthesis tree but not the taxonomy tree. For
example, in Figure \ref{fig:Handling-incertae-sedis}, taxon $C$ is a
descendant of taxon $A$ in the synthetic tree (Fig. \ref{fig:Handling-
incertae-sedis}d), but not the taxonomy (Fig. \ref{fig:Handling-
incertae-sedis}b). Thus, we could say that the synthetic tree places
$C$ within $A$. Furthermore, the synthetic tree additionally places
$C$ within $AB$.

A slightly more general case is shown in Figure \ref{fig:Broken-
incertae-sedis}. Here taxon $C1$ is an descendant of $A$ in the
synthesis tree, but not the taxonomy. Thus $C1$ is placed within $A$.
We note that the most recent common ancestor (MRCA) of $A$ and $C1$ in
the taxonomy is the root node, and the path to the MRCA from $A$ is
$A\to AB\to root$, while the path to the MRCA from $C1$ is $C1\to C\to
root$. Here the additional node $AB$ on the path to the MRCA from $A$
indicates that $C1$ is placed within $AB$ as well as $A$. In contrast,
the additional node $C$ on the path to the MRCA indicates that $C1$ is
a broken \emph{incertae sedis} taxon. Thus we may consider an
algorithm that, for each taxon $B$ on the synthesis tree, finds the
closest ancestral taxon $A$, and checks if $A$ is an ancestor of $B$
on the taxonomy. If not, we compute the paths $A\to A_{1}\to\ldots\to
MRCA(A,B)$ and $B\to B_{1}\to\ldots\to MRCA(A,B)$.

However, this assumes that \emph{incertae sedis} taxa are always
placed within their siblings, or descendants of their siblings. As
such, it does not handle cases such as Figure \ref{fig:Placing-one-
incertae}, in which $I$ is first placed as sibling to $AI$, and then
$AI$ is placed within $I$. Note that $I$ is not a sibling of $AI$ or a
descendant of a sibling. Furthermore, the MRCA of $I$ and $AI$ is the
root, and the path from $AI$ to the root includes $A$, but $A$ is not
broken. We suggest that this conundrum may be resolved by first
processing the placement of $I$ within $A$ by modifying the taxonomy
to reflect this placement\footnote{Does this actually work? Some
placements break taxa.}. When we consider the placement of $AI$ within
$I$ on this modified taxonomy, the MRCA of $I$ and $AI$ is now $A$
instead of the root, and the correct interpretation results. We
suggest that by walking the tree in a pre-order fashion and processing
shallowing placements first, the previous approach of finding the MRCA
and labelling intermediate nodes as either additional placements or
broken incertae sedis (or incertae\_sedis\_inherited) taxa holds. Such
an approach should also handle cases containing nested incertae sedis
taxa, as in Figure \ref{fig:Placing-one-incertae}.

\begin{comment}
One question is whether this approach is able to handle nested
incertae sedis taxa.

@@

Each input tree can relate to an \emph{incertae sedis} taxon
$(A,B,C)D$ in a number of ways
\begin{itemize}
    \item it could resolve
$A$, $B$, $C$, or $D$.
    \item it could place $D$ on a degree-2 (=out-
degree-1) node that bisects a branch 
    \item it could place a descendant
taxon of $D$ 
    \item it could place a descendant taxon of $D$ in a
\emph{different place} than another input tree.
    \item it could place
children of $D$ in multiple places, thus conflicting with the branch.
If the \emph{incertae sedis} taxon $(A,B,C)$ is broken, then $A$, $B$
and $C$ become \emph{incertae sedis} clades in their own right, that
may attach separately, except that they . This is because none of $A$,
$B$, or $C$ is in the exclude set of the siblings of $D$.
\end{itemize}
\end{comment}

@@

When the synthesis tree conflicts with a taxonomy node, we say that
the taxon $B$ at that node is a broken taxon. Broken taxa have two
main effects, both of which are negative. First, the name $B$ of the
broken taxon is removed from the synthesis tree. Second, the
conflicting edge for taxon $B$ is not included in the synthesis tree.
This means that any children of $B$ that are taxonomy-only will not be
placed with the children of $B$ that are mentioned in input
phylogenies. Instead the taxonomy-only children of a broken taxon move
towards the root of the synthesis tree and attach at the first higher-
ranked taxon that is an ancestor of $B$ but is not broken.

We therefore seek a synthesis method that can correctly introduce
\emph{incertae sedis} taxa into containing taxa without breaking the
containing taxa. We change the semantics of names to imply, not the
exclusion of all non-included taxa, but only some non-included taxa,
as described in section \ref{sec:Semantics-of- incertae}. As a result
of this change in semantics, placing an IS taxon $A$ within a sister
taxon $B$ no longer results in conflict with $B$. This allows us to
retain the split for $B$ within the synthesis tree, so that taxonomy-
only children of $B$ are correctly grouped with their siblings that
are referenced by the input trees. We may then retain the name for the
no-longer-broken taxon. Finally, we are then able to stop filtering
\emph{incertae sedis} taxa, so that they appear in the synthesis tree.
Thus, the synthesis tree is able to represent substantially more
species, without suffering the loss of taxa and the loss of structure.

@@

{[}\textbf{BDR:} \emph{maybe we should merge this with Figure.
\ref{fig:Broken- incertae-sedis}. Although the point of figure
\ref{fig:placement-example1} is to talk about sub-problem
decomposition as well as what the result looks like.}{]}

The addition of \emph{incertae sedis} taxa allows new types of
conflict between input trees. For example, different input trees might
place an incertae sedis taxon in conflicting locations. This is
illustrated in Figure \ref{fig :placement-example1}, where the IS
taxon (ott7,ott6)ott10 is placed as sister to ott1 by phylogeny
$\tau_{1}$ and as sister ott3 by phylogeny $\tau_{2}$.

@@

\subsection{Case 1: An emph{incertae sedis} clade}

When a clade is marked as \emph{incertae sedis}, we need to consider
two cases. In the first case, the clade may be placed within a sister
clade intact (Figure \ref{fig:Handling-incertae-sedis}). However, if
the clade is not monophyletic in the synthesis tree, then we allow the
members of the clade to be placed separately (Figure \ref{fig:Broken-
incertae-sedis}).

@@

\end{figure}
\begin{figure}
\hfill{}\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/tree1-shaded}

@@

\subsection{Case 2: A nested emph{incertae sedis} clade}

@@

\begin{figure}
\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}

@@

\section{Handling emph{incertae sedis} taxa in the propinquity
pipeline}

@@

One current problem is that well-known taxa like Fungi or Mammalia
tend to have a very large number of incertae sedis children, making
browsing in the tree viewer difficult. This can happen when, for
example, fossils or other hard-to- place taxa get classified only to
the level of these well-known nodes and no further. This leads to a
situation where well-known taxa serve as a dumping ground for unplaced
taxa.

@@

The presence of \emph{incertae sedis }taxa poses a problem to sub-
problem decomposition, since taxonomy edges no longer completely
separate subproblems. Instead, \emph{incertae sedis} taxa may attach
on either side of a taxonomy edge. We seek to place \emph{incertae
sedis} taxa into subproblems in such a way that the subproblem solver
can perform the placement inside the subproblem. This approach
postpones handling of conflict in \emph{incertae sedis} taxa to the
subproblem solver, where the problem is well formulated in terms of
splits. However, it does have the effect of creating larger
subproblems.
